---
title: "Project"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Project: Yacht Hydrodynamics
Trying out linear regression on this dataset.
TODO:
1. Check code and move commentary out out of code section.
2. Add details on how the two models differ and why one model could be better to use than the other.
3. Add info on how informative priors were calculated and used but it actually turned out to have worse variance for the response variable - Seem to be getting negative R^2 value when using informative priors - LOOK INTO IT!
4. Write commentary on how we chose the 1st model since it's DIF was the lowest and then we went on to use the model to predict values in test dataset and plot the model parameters with predicted values (and with upper/lower credible sets)
5. Understand more around what happens after model is created to plot everything. This link was used as reference: http://biometry.github.io/APES//LectureNotes/StatsCafe/Linear_models_jags.html
6. Structure for report: Split code into sections and write commentary for each section followed by corresponding charts. 
7. Figure out why there are negative values being predicted 
8. Figure out what's the difference between MCMC that JAGS uses vs what gets created in the last few sections of first model

More things to try out in the modeling side of the report:
5. Find model performance indicators such as RMSE using predicted values in test dataset


## Model 1: with all parameters

```{r Project: model 1, echo=TRUE}
#Include necessary package
library(R2jags)
library(rjags)
library('corrplot')
library('PerformanceAnalytics')

# set the seed to make partitioning of training vs set reproducible
set.seed(12345)

# Upload data into dataframes
yacht_data <- read.table("yacht_hydrodynamics.data", header=FALSE)
names(yacht_data) <- c(
  "longitudinal_position_buoyancy",
  "prismatic_coeff",
  "length_displacement_ratio",
  "beam_draught_ratio",
  "length_beam_ratio",
  "froude_number",
  "residuary_resistance"
  )

# Descriptive statistics
summary(yacht_data)

# Feature engineering
# Evaluating correlations
# Correlation matrix table
res <- cor(yacht_data)
round(res, 2)

# Correlation matrix chart
corrplot(as.matrix(cor(yacht_data)))

# Splitting between train and test for cross-validation of model
# Dedicating 80% and 20% of dataset for train and test respectively
# 80% of the sample size
sample_size <- floor(0.80 * nrow(yacht_data))
train_ind <- sample(seq_len(nrow(yacht_data)), size = sample_size)
train_yacht_data <- yacht_data[train_ind, ]
test_yacht_data <- yacht_data[-train_ind, ]

# Looking at correlation matrix, it looks like length_displacement_ratio and length_beam_ratio have high correlation (0.68), hence ignoring one of the variables at random.
# 1. But before that, trying out linear regression with all predictors included.

# define data structure required for JAGS
yacht_jags_data <- with(train_yacht_data, list(
    longitudinal_position_buoyancy = longitudinal_position_buoyancy, 
    prismatic_coeff = prismatic_coeff,
    length_displacement_ratio = length_displacement_ratio,
    beam_draught_ratio = beam_draught_ratio,
    length_beam_ratio = length_beam_ratio,
    froude_number = froude_number,
    residuary_resistance = residuary_resistance,
    N_train = nrow(train_yacht_data)
    )
  )

yacht_jags_data_test <- with(test_yacht_data, list(
    longitudinal_position_buoyancy = longitudinal_position_buoyancy, 
    prismatic_coeff = prismatic_coeff,
    length_displacement_ratio = length_displacement_ratio,
    beam_draught_ratio = beam_draught_ratio,
    length_beam_ratio = length_beam_ratio,
    froude_number = froude_number,
    residuary_resistance = residuary_resistance,
    N_test = nrow(test_yacht_data)
    )
  )

linear_reg_model_jags <- function(){

  # Likelihood
	for (i in 1:N_train){
		mu[i] <- beta0 + beta1 * longitudinal_position_buoyancy[i] + beta2 * prismatic_coeff[i] + beta3 * length_displacement_ratio[i] + beta4 * beam_draught_ratio[i] + beta5 * length_beam_ratio[i] + beta6 * froude_number[i]
		residuary_resistance[i] ~ dnorm(mu[i], tau) 
	}

  # Priors
	beta0 ~ dnorm(0, 0.01) # intercept
	beta1 ~ dnorm(0, 0.01) # slopes
	beta2 ~ dnorm(0, 0.01)
	beta3 ~ dnorm(0, 0.01) 
	beta4 ~ dnorm(0, 0.01)
	beta5 ~ dnorm(0, 0.01)
	beta6 ~ dnorm(0, 0.01)
	sigma ~ dunif(0, 100) # SD
	tau <- 1 / (sigma * sigma)

}

# initialize values for params
init_values <- function(){
	list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), beta3 = rnorm(1), beta4 = rnorm(1), beta5 = rnorm(1), beta6 = rnorm(1), sigma = runif(1))
}
params <- c("beta0", "beta1", "beta2", "beta3", "beta4", "beta5", "beta6", "sigma")

# Fit the model
fit_linear_reg_model <- jags(data = yacht_jags_data, inits = init_values, parameters.to.save = params, model.file = linear_reg_model_jags,
			   n.chains = 3, n.iter = 12000, n.burnin = 2000, n.thin = 10, DIC = TRUE)

# View the results of the model
fit_linear_reg_model

```

## Model 2: Now trying out by ignoring length_beam_ratio

```{r model 2, echo=TRUE}

# define data structure required for JAGS
yacht_jags_data_multi_coll <- with(train_yacht_data, list(
    longitudinal_position_buoyancy = longitudinal_position_buoyancy, 
    prismatic_coeff = prismatic_coeff,
    length_displacement_ratio = length_displacement_ratio,
    beam_draught_ratio = beam_draught_ratio,
    froude_number = froude_number,
    residuary_resistance = residuary_resistance,
    N_train = length(residuary_resistance)
    )
  )

linear_reg_model_jags_multi_coll <- function(){
	# Likelihood:
	for (i in 1:N_train){
		residuary_resistance[i] ~ dnorm(mu[i], tau) # tau is precision (1 / variance)
		mu[i] <- beta0 + beta1 * longitudinal_position_buoyancy[i] + beta2 * prismatic_coeff[i] + beta3 * length_displacement_ratio[i] + beta4 * beam_draught_ratio[i] + beta5 * froude_number[i]
	}
	# Priors:
	beta0 ~ dnorm(0, 0.01) # intercept
	beta1 ~ dnorm(0, 0.01) # slopes
	beta2 ~ dnorm(0, 0.01)
	beta3 ~ dnorm(0, 0.01) 
	beta4 ~ dnorm(0, 0.01)
	beta5 ~ dnorm(0, 0.01)
	sigma ~ dunif(0, 100) # standard deviation
	tau <- 1 / (sigma * sigma) # sigma^2 doesn't work in JAGS
	
}

# initialize values for params
init_values <- function(){
	list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), beta3 = rnorm(1), beta4 = rnorm(1), beta5 = rnorm(1), sigma = runif(1))
}
params <- c("beta0", "beta1", "beta2", "beta3", "beta4", "beta5", "sigma")

# Fit the model
fit_linear_reg_model_multi_coll <- jags(data = yacht_jags_data_multi_coll, inits = init_values, parameters.to.save = params, model.file = linear_reg_model_jags_multi_coll,
			   n.chains = 3, n.iter = 12000, n.burnin = 2000, n.thin = 10, DIC = TRUE)

# View the results of the model
fit_linear_reg_model_multi_coll
```

## Model 3. Trying out with user-specified Priors instead of non-informative priors. 

```{r model 3, echo=TRUE}
# 3. Trying out with user-specified Priors instead of non-informative priors. 
# Using average and precision of sample observations (dataset) as prior mean and precision
linear_reg_model_jags <- function(){
	# Likelihood:
	for (i in 1:N_train){
		residuary_resistance[i] ~ dnorm(mu[i], tau) # tau is precision (1 / variance)
		mu[i] <- beta0 + beta1 * longitudinal_position_buoyancy[i] + beta2 * prismatic_coeff[i] + beta3 * length_displacement_ratio[i] + beta4 * beam_draught_ratio[i] + beta5 * length_beam_ratio[i] + beta6 * froude_number[i]
	}
	# Priors:
	beta0 ~ dnorm(0, 0.01) # intercept
	beta1 ~ dnorm(0, 0.4) # slopes
	beta2 ~ dnorm(0.5, 1000)
	beta3 ~ dnorm(4.7, 15) 
	beta4 ~ dnorm(3.9, 3)
	beta5 ~ dnorm(3.2, 16)
	beta6 ~ dnorm(0.2, 98)
	sigma ~ dunif(0, 100) # standard deviation
	tau <- 1 / (sigma * sigma) # sigma^2 doesn't work in JAGS
	
}

# initialize values for params
init_values <- function(){
	list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), beta3 = rnorm(1), beta4 = rnorm(1), beta5 = rnorm(1), beta6 = rnorm(1), sigma = runif(1))
}
params <- c("beta0", "beta1", "beta2", "beta3", "beta4", "beta5", "beta6", "sigma")

# Fit the model
fit_linear_reg_model_informative_prior <- jags(data = yacht_jags_data, inits = init_values, parameters.to.save = params, model.file = linear_reg_model_jags,
			   n.chains = 3, n.iter = 12000, n.burnin = 2000, n.thin = 10, DIC = TRUE)

# View the results of the model
fit_linear_reg_model_informative_prior
```

```{r Proceeding with further steps on Model 2 (excluding 1 predictor), echo=TRUE}

fit_linear_reg_model <- fit_linear_reg_model_multi_coll
# Visual check of how the MCMC chains have mixed
traceplot(fit_linear_reg_model, mfrow = c(2, 2), ask = F)

# View the parameters' marginal posterior distribution 
plot(fit_linear_reg_model)

# Plot the model
fit_linear_reg_model_mcmc <- as.mcmc(fit_linear_reg_model)

# Using the model to predict residuary resistance for the test dataset
# Combining the instances of MCMC first
fit_linear_reg_model_mcmc_merge <- as.mcmc(rbind(fit_linear_reg_model_mcmc[[1]], 
                                fit_linear_reg_model_mcmc[[2]], 
                                fit_linear_reg_model_mcmc[[3]]))

# Estimating predicted value of residuary resistance for each predictor value in test dataset 
prediction_mean <- mean(fit_linear_reg_model_mcmc_merge[, "beta0"]) + mean(fit_linear_reg_model_mcmc_merge[, "beta1"]) * yacht_jags_data_test$longitudinal_position_buoyancy + mean(fit_linear_reg_model_mcmc_merge[, "beta2"]) * yacht_jags_data_test$prismatic_coeff +  mean(fit_linear_reg_model_mcmc_merge[, "beta3"]) * yacht_jags_data_test$length_displacement_ratio + mean(fit_linear_reg_model_mcmc_merge[, "beta4"]) * yacht_jags_data_test$beam_draught_ratio + mean(fit_linear_reg_model_mcmc_merge[, "beta5"]) * yacht_jags_data_test$froude_number

prediction_mean_distribution <- matrix(NA, nrow = nrow(fit_linear_reg_model_mcmc_merge), ncol = yacht_jags_data_test$N_test)
for (i in 1:nrow(prediction_mean_distribution)){
	prediction_mean_distribution[i,] <- fit_linear_reg_model_mcmc_merge[i,"beta0"] + fit_linear_reg_model_mcmc_merge[i, "beta1"] * yacht_jags_data_test$longitudinal_position_buoyancy + fit_linear_reg_model_mcmc_merge[i, "beta2"] * yacht_jags_data_test$prismatic_coeff +  fit_linear_reg_model_mcmc_merge[i, "beta3"] * yacht_jags_data_test$length_displacement_ratio + fit_linear_reg_model_mcmc_merge[i, "beta4"] * yacht_jags_data_test$beam_draught_ratio + fit_linear_reg_model_mcmc_merge[i, "beta5"] * yacht_jags_data_test$froude_number
}
credible_lower <- apply(prediction_mean_distribution, MARGIN = 2, quantile, prob = 0.025)
credible_upper <- apply(prediction_mean_distribution, MARGIN = 2, quantile, prob = 0.975)

# Plot mean model prediction and the lower/upper credible sets
# Choosing only froude_number to visualize since it has the highest correlation with response var
plot(yacht_jags_data$residuary_resistance ~ yacht_jags_data$froude_number, data = yacht_jags_data)
lines(yacht_jags_data_test$froude_number, prediction_mean)
lines(yacht_jags_data_test$froude_number, credible_lower, lty = 2, col = "red")
lines(yacht_jags_data_test$froude_number, credible_upper, lty = 2, col = "red")
```